---
title: "Data Analysis on Volcano Eruptions and Environmental Impact"
format: revealjs
editor: visual
---

```{r setup, echo: false, message: false, warning: false, show_col_types = FALSE}
# Load necessary libraries
library("tidyverse")
library("ggplot2")
library("dplyr")
library("ggpubr")

# Load datasets 
eruptions <- read_csv("tidytuesday:2020:2020-05-12:volcano/eruptions.csv", show_col_types = FALSE)
env_impact <- read_csv("tidytuesday:2020:2020-05-12:volcano/tree_rings.csv", show_col_types = FALSE)

```

---

## Project Overview

This study explores the relationship between volcanic eruptions, represented by the Volcanic Explosivity Index (VEI), and their environmental impact. The data is provided by TidyTuesday.

---

## Data Overview

The analysis uses the following datasets and key variables:

-   **Eruptions:** Contains eruption years, locations (latitude, longitude), and VEI values (Volcanic Explosivity Index (0 to 8 scale))
-   **Env_impact:** Records environmental data, including the Europe Temperature Index (°C, relative to 1961–1990) and n_tree (Tree ring z-scores (relative to 1000–1099))

## ![VEI-Wikipedia](https://upload.wikimedia.org/wikipedia/commons/0/01/VEIfigure_en.svg)

---
## Mapping: Spatial Distribution of VEI

The first analysis visualizes the spatial distribution of volcanic eruptions by VEI.

```{r}
# Mapping VEI values across the globe
ggplot(data = eruptions %>% filter(!is.na(vei)), aes(x = longitude, y = latitude, color = vei)) +
  borders("world", colour = "gray80", fill = "gray95") +
  geom_point(size = 2, alpha = 0.5) +
  scale_color_gradient(low = "blue", high = "red") +
  theme_minimal() +
  labs(
    title = "Map of Volcano Eruptions with VEI Scale",
    x = "Longitude",
    y = "Latitude",
    color = "VEI"
  )
```

---

## Correlation: Tree ring z-score vs. Europe Temperature Index

Next, we analyze the relationship between the Tree ring z-scores and the Europe Temperature Index.

```{r}
# Correlation analysis
ggplot(env_impact, aes(x = n_tree, y = europe_temp_index)) + 
  geom_point(size = 1, stroke = 0.8, color = "grey40") + 
  geom_smooth(method = "lm", color = "red") + 
  stat_cor(aes(label = ..r.label..), method = "pearson", color = "black", size = 4) +
  ggtitle("Correlation Between Tree Ring Z-Scores and Europe Temperature Index") +
  theme_minimal() +
  xlab("Tree Ring Z-Scores") +
  ylab("Europe Temperature Index")
```

---