---
title: "Data Analysis on Volcano Eruptions and Environmental Impact"
author: "`Jeongyeon Yun (Jennie)`"
format: revealjs
editor: visual
---

```{r setup, echo: false, message: false, warning: false, show_col_types = FALSE}
#| echo: FALSE

# Load relevant library
library(dplyr) 
library(tidyr)
library(ggplot2) 
library(ggpubr)
library(readr)
library(gridExtra)
library(patchwork)

# Load datasets 
eruptions <- read_csv("tidytuesday:2020:2020-05-12:volcano/eruptions.csv", show_col_types = FALSE)
env_impact <- read_csv("tidytuesday:2020:2020-05-12:volcano/tree_rings.csv", show_col_types = FALSE)

```

------------------------------------------------------------------------

## Project Overview

This project explores the relationship between volcanic eruptions, represented by the Volcanic Explosivity Index (VEI), and their environmental impact.

The dataset is sourced from TidyTuesday.

![VEI-Wikipedia](https://upload.wikimedia.org/wikipedia/commons/0/01/VEIfigure_en.svg)

::: notes
In this presentation, I'll guide you through how I used different visualization techniques to explore the relationship between volcanic eruptions, described as their Volcanic Explosivity Index (VEI), and environmental impacts. For you to make it sense what I'm explaining, inevitably I need to explain somewhat about my dataset. But mainly I will focus on how I used various methods to represent the data effectively.
:::

------------------------------------------------------------------------

## Data Overview

The analysis uses the following datasets and key variables:

**Eruptions**

Variables: Eruption year, location (latitude, longitude), VEI (scale: 0 to 8). Environmental Impact:

**Key Variables**

Europe Temperature Index (°C, relative to 1961–1990 baseline). Tree ring z-scores (relative to 1000–1099)

--- 

## Dataset quality 
The dataset appears to be organized in a tidy format, which facilitates analysis. 
However, due to differences in observation periods and units across the datasets, 
some data preprocessing was necessary to ensure compatibility for time series 

--- 

## [Mapping: Spatial Distribution of volcano eruptions by VEI]{.r-fit-text}

-   The map highlights regions with varying levels of volcanic intensity, with a color-gradient representing different VEI values

```{r, echo=TRUE}
#| code-fold: true

# Creating a map to show the spatial distribution of volcano eruptions by VEI
ggplot(data = eruptions %>% filter(!is.na(vei)), aes(x = longitude, y = latitude, color = vei)) +
  borders("world", colour = "gray80", fill = "gray95") +  # Add a world map with gray borders and light background
  geom_point(size = 2, alpha = 0.5) +  # Plot eruption points 
  scale_color_gradient(low = "blue", high = "red") +  # Color scale: Blue for low VEI, red for high VEI
  theme_minimal() +  # Minimalist theme for a clean look
  labs(title = "Map of Volcano Eruptions with VEI Scale",  # Title of the plot
       x = "Longitude",  # Label for the x-axis
       y = "Latitude",  # Label for the y-axis
       color = "VEI") +  # Label for the color scale
  theme(plot.title = element_text(size = 30, face = "bold")) # Increase title size for better visibility
```

::: notes
For the first visualization, I used a map to represent the spatial distribution of volcanic eruptions based on their VEI values. I chose a world map as the base, with eruption locations marked by colored points. The colors range from blue for smaller eruptions to red for larger ones, which helps quickly communicate the distribution of eruption sizes across the globe.
I used ggplot2 in R, with the geom_point() function for the points and borders() to map the world. By applying a gradient color scale, it makes it easy to understand the global patterns of volcanic activity
:::
---
## [Correlation: Tree ring z-score vs. Europe Temperature Index]{.r-fit-text}

- Visualize the relationship between the Tree ring z-scores and the Europe Temperature Index

```{r, echo=TRUE}
#| code-fold: true
ggplot(env_impact, aes(x = n_tree, y = europe_temp_index)) + 
  geom_point(size = 1, stroke = 0.8, color = "grey40") + 
  geom_smooth(method = "lm", color = "red") + 
  stat_cor(aes(label = ..r.label..), method = "pearson", color = "black", size = 7) +
  ggtitle("Correlation Between Tree Ring Z-Scores and Europe Temperature Index") +
  theme_minimal() +
  xlab("Tree Ring Z-Scores") +
  ylab("Europe Temperature Index")+
  theme(
    plot.title = element_text(size = 30, face = "bold")  # Increase title size
  )
```
:::notes
Next, I visualized the relationship between the tree ring z-scores and the Europe Temperature Index using a scatter plot. This is important because it helps us understand how volcanic activity could affect long-term environmental changes.
I utilized geom_point() for individual data points, and added a linear regression line using geom_smooth() to show the trend. To quantify the correlation, I applied stat_cor(), which automatically calculates and displays the Pearson correlation coefficient, giving us a numeric measure of the strength of the relationship
:::
---
## Multiplots: Comparing the Impact of VEI on Europe Temperature Over Time{.r-fit-text}

- Visualize the data in two separate plots to investigate how volcanic eruptions categorized by their Volcanic Explosivity Index (VEI), impact the Europe Temperature Index (ETI) in the same year  

---

```{r}
# Merge env_impact data with eruption data
env_impact_vei <- inner_join(env_impact, eruptions, by = c("year" = "start_year"))

# Categorize VEI into levels (1-2, 3-4, >=5)
env_impact_vei <- env_impact_vei %>%
  mutate(vei_category = case_when(
    vei >= 5 ~ "VEI >= 5",  # Large eruptions
    vei >= 3 ~ "VEI 3-4",    # Moderate eruptions
    vei >= 1 ~ "VEI 1-2",    # Small eruptions
    TRUE ~ NA_character_    
  ))

# Calculate average Europe Temperature Index for each VEI category
vei_temp_index <- env_impact_vei %>%
  filter(!is.na(vei_category)) %>%
  group_by(vei_category) %>%
  summarise(avg_temp_index = mean(europe_temp_index, na.rm = TRUE)) %>%
  mutate(vei_category = factor(vei_category, levels = c("VEI 1-2", "VEI 3-4", "VEI >= 5")))

# Plot 1: VEI Category vs Average Europe Temperature Index
plot1 <- ggplot(vei_temp_index, aes(x = vei_category, y = avg_temp_index, fill = vei_category)) +
  geom_bar(stat = "identity") + 
  geom_text(aes(label = round(avg_temp_index, 2)), vjust = -0.5) +  # Show average on top of bars
  labs(x = "VEI Category", y = "Average ETI") +
  theme_minimal() + 
  theme(
    legend.position = "top"  # Move legend to the top
  )

# Plot 2: VEI Intensity and Europe Temperature Index Over Time
# Calculate VEI intensity by year
vei_intensity <- env_impact_vei %>%
  group_by(year, vei_category) %>%
  summarise(vei_count = n()) %>%
  mutate(vei_intensity_value = case_when(
    vei_category == "VEI >= 5" ~ vei_count * 10000,
    vei_category == "VEI 3-4" ~ vei_count * 100,
    vei_category == "VEI 1-2" ~ vei_count * 1,
    TRUE ~ 0
  ))

# Total VEI intensity per year
vei_intensity_total <- vei_intensity %>%
  group_by(year) %>%
  summarise(total_intensity = sum(vei_intensity_value))

# Join VEI intensity data with Europe Temperature Index data
vei_temp_joined <- left_join(vei_intensity_total, env_impact_vei, by = "year")

# Plot 2: Combine bar chart (VEI intensity) and line chart (Temperature Index)
plot2 <- ggplot(vei_temp_joined, aes(x = year)) +
  geom_bar(aes(y = total_intensity, fill = "VEI Intensity"), stat = "identity", alpha = 0.9, linewidth = 10) +
  geom_line(aes(y = europe_temp_index * 100000, color = "Europe Temperature Index"), size = 1) +
  scale_y_continuous(
    name = "VEI Intensity",
    sec.axis = sec_axis(~ . / 100000, name = "Europe Temperature Index (°C)")
  ) +
  scale_fill_manual(values = c("VEI Intensity" = "blue")) +
  scale_color_manual(values = c("Europe Temperature Index" = "red")) +
  labs(x = "Year") +
  theme_minimal()+
  theme(legend.position = "top")

# Combine the plots with adjusted spacing
plot1 + plot2 + 
  plot_annotation(tag_levels = 'a',
                title = "VEI intensity effect on Europe temperature",
                subtitle = "a:Average ETI by VEI category b:VEI and ETI over time")+
  plot_layout(widths = c(3, 11)) +
  theme(legend.position = "top", 
        legend.title = element_text(size =10),
        legend.text = element_text(size = 10),
        legend.key.size = unit(0.2, 'cm'),
        plot.tag = element_text(size = 11, face = 'bold'),
        plot.title = element_text(size = 25, face = 'bold'),  # Increase title size
        plot.subtitle = element_text(size = 20)) 
```
:::notes
For the third set of visualizations, I aimed to compare the impact of different VEI categories on the Europe Temperature Index over time.
In the first plot, I grouped eruptions by their VEI categories, calculated the average ETI for each group, and represented this data in a bar chart. The bars show how large, moderate, and small eruptions affect the temperature, with the text labels on top providing the exact values for clarity.
For the second plot, I visualized both VEI intensity over time and the ETI in a dual-axis plot. This allows us to compare the number of eruptions and their intensity directly against changes in the Europe Temperature Index. The bars represent VEI intensity, while the line shows the temperature index, making the relationship between the two more obvious
:::
---
```{r, echo=TRUE}
# Plot 2: VEI Intensity and Europe Temperature Index Over Time
# Calculate VEI intensity by year
vei_intensity <- env_impact_vei %>%
  group_by(year, vei_category) %>%
  summarise(vei_count = n()) %>%
  mutate(vei_intensity_value = case_when(
    vei_category == "VEI >= 5" ~ vei_count * 10000,
    vei_category == "VEI 3-4" ~ vei_count * 100,
    vei_category == "VEI 1-2" ~ vei_count * 1,
    TRUE ~ 0
  ))
```
---